---
title: "annotate_450k_X_CpGs"
author: "Yunfeng"
date: "1/16/2023"
output: html_document
---
#Script information

This script was written to annotate X-chromosome CpGs of the Illumina Infinium 450K array with CGI features and XCI status.

#CGI feature annotation

CGI-centric annotation was split in 3 regions:
1)	CGIs, as annotated by UCSC (hg19).
2)	Shores: 2 kb regions flanking the CGIs (both up- and downstream).
3)	Non-CGI: none of the above.

Schematic representation of the CGI-centric annotation:

----------------------|-------|-------|-------|----------------------
       Non-CGI        | Shore |  CGI  | Shore |        Non-CGI
                      | 2 kb  |       | 2 kb  |
----------------------|-------|-------|-------|----------------------

# XCI status annotation
XCI status annotation were based on reliable meta-status calls[1], which integrated three published studies and resulted in genes classified into three categories: subject to XCI, escape XCI and variably escape XCI.

[1] Balaton BP, Brown CJ: Contribution of genetic and epigenetic changes to escape from X-chromosome inactivation. Epigenetics Chromatin 2021, 14:30.

```{r}
#set library path.
.libPaths("~/researchdrive/yliu/Rlibs")
```


```{r}
#Load all necessary libraries.
library(EnsDb.Hsapiens.v75)
library(rtracklayer)
library(FDb.InfiniumMethylation.hg19)
library(AnnotationHub)
library(ph525x)
library(data.table)
```

```{r}
#Load data.
load(file ="./Output/Output_01_chrX.RData")

#Make a GRanges object containing features of all 450K CpGs.
ann450K <- features(FDb.InfiniumMethylation.hg19)

#Select only the X chromosome CpGs which were measured in the BIOS data.
mvalues_chrX<-cbind(mvalues_Xfemales,mvalues_Xmales)
dim(mvalues_chrX) # 9777 4031

ann450K_chrX <- ann450K[rownames(mvalues_chrX)]

#Remove  metadata and strand information, then sort the object.
ann450K_chrX <- ann450K_chrX[,NULL]
strand(ann450K_chrX) <- "*"
ann450K_chrX <- sort(sortSeqlevels(ann450K_chrX))

#Add the chromosome names to the metadata.
ann450K_chrX$Chromosome <- factor(seqnames(ann450K_chrX))
ann450K_chrX
```

```{r}
# Open ensembl database.
edb75 <- EnsDb.Hsapiens.v75
# Only keep protein-coding genes.
GeneEns <- genes(edb75, filter = GeneBiotypeFilter("protein_coding"))
GeneEns <- keepStandardChromosomes(GeneEns, pruning.mode = "coarse")
GeneEns <- sort(sortSeqlevels(GeneEns))
values(GeneEns)[,c(2,4)] <- NULL
GeneEns

# keep protein-coding genes on X chromosome
seqlevels(GeneEns) <- gsub("X", "chrX", seqlevels(GeneEns))
GeneEns_chrX <- keepSeqlevels(GeneEns,"chrX", pruning.mode = "coarse")
```

```{r}
#Change the genome nomenclature of GeneEns_chrX from GRCh37 to hg19
genome(GeneEns_chrX)<-genome(ann450K_chrX) 
ann450K_chrX
GeneEns_chrX
```


```{r}
#Annotate CGIs
#Retrieve CpG islands from the UCSC genome browser.
mySession = browserSession("UCSC")
genome(mySession) <- "hg19"
CpGislands.raw <- getTable(ucscTableQuery(mySession, track="CpG Islands",table="cpgIslandExt"))
CGI.gr <- GRanges(CpGislands.raw$chrom, IRanges(CpGislands.raw$chromStart+1, CpGislands.raw$chromEnd), name=CpGislands.raw$name)
CGI.gr <- keepStandardChromosomes(CGI.gr, pruning.mode = "coarse")

#Add CGI annotation (also including shores (within 2000 bp from CGI)).
#Shores start
shores1 <- CGI.gr
shores2 <- CGI.gr
start(shores1) <- start(CGI.gr) - 2000
end(shores1) <- start(CGI.gr)

#Shores end
end(shores2) <- end(CGI.gr) + 2000
start(shores2) <- end(CGI.gr) 
shores <- c(shores1, shores2)

#Add the CGI annotations to the 450K CpGs.

#Non-CGI
ann450K_chrX$CGI_Feature <- rep("non-CGI")

#Shore
OL.sho <- as.matrix(findOverlaps(shores,ann450K_chrX))[,2]
OL.sho <- unique(OL.sho)
ann450K_chrX[OL.sho,]$CGI_Feature <- "Shore"

#CGI
OL.cgi <- as.matrix(findOverlaps(CGI.gr, ann450K_chrX))[,2]
OL.cgi <- unique(OL.cgi)
ann450K_chrX[OL.cgi,]$CGI_Feature <- "CGI"

#Turn CGI features into a factor variable.
ann450K_chrX$CGI_Feature <- factor(ann450K_chrX$CGI_Feature, levels = c("CGI", "Shore", "non-CGI"))

ann450K_chrX
```


```{r}
# XCI annotation based on meta calls
load(file="Input_07_TSS.XCI.Status.meta.calls.RData")
XCI.annotation<-Annotation.TSS.meta.calls[names(ann450K_chrX),]
ann450K_chrX$XCI.Status.meta.calls<-XCI.annotation$TSS.Status
ann450K_chrX$Distance.meta.calls<-XCI.annotation$Distance
```


```{r}
#save the files.
save(GeneEns_chrX, file = "Output/Output_07/Output_07_ensembl_genes_chrX.RData")
save(ann450K_chrX, file = "Output/Output_07/Output_07_450K_chrX_annotation.RData")
```
